---
title: "ModelSelection"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
```

# Added - Will later remove

Added this part to make the dataset and tree. Same as in Pipeline.Rmd.
Phyloglmstep starts at the header Model Selection

#### data and tree
```{r declare root and paths}
root <- "/home/zoe/Documents/GitHub/trait-organismal-ungulates-R"
treeFile <- paste(c(root, "/data/ungulates.tree"), collapse = "")
ungulatesFile <- paste(c(root, "/data/CSV/ungulatesTraits.csv"), collapse = "")
omiFile <- ("https://raw.githubusercontent.com/naturalis/trait-geo-diverse-ungulates/master/results/OMI/niche_traits.csv")

ungulatesData <- read.csv(ungulatesFile, header = TRUE, sep= ",")
omi <- read.csv(omiFile)
tree <- read.tree(treeFile)

ungulatesData <- ungulatesData[2:60]
omi <- omi %>% rename(CanonicalName = X)
dataset <- merge(ungulatesData, omi, by="CanonicalName")
dataset$CanonicalName <- gsub(" ", "_", dataset$CanonicalName)

# Remove all unnecessary variables 
rm(treeFile, ungulatesFile, omiFile, ungulatesData, omi)

# Which species aren't in the tree
dropRows <- setdiff(dataset$CanonicalName, tree$tip.label)

# Drop rows that aren't in the tree, and aren't domesticated
row.names(dataset) <- dataset$CanonicalName
dataset <- dataset[!(row.names(dataset) %in% dropRows), ]

# Drop tips that aren't in dataset
dropTips <- setdiff(tree$tip.label, dataset$CanonicalName)
tree <- drop.tip(tree, dropTips)

rm(dropRows, dropTips)

#rename columns with dots in the name
dataset <- dataset %>% rename(Horns_Antlers = Horns.Antlers,
                              X21.1_PopulationDensity_n_km2 = X21.1_PopulationDensity_n.km2)


#remove traits that (almost) only consist of missing values (>100 NA)
dataset <- subset(dataset, select = -c(X5.4_WeaningBodyMass_g, X13.3_WeaningHeadBodyLen_mm, X13.2_NeonateHeadBodyLen_mm, X18.1_BasalMetRate_mLO2hr, X5.2_BasalMetRateMass_g, X2.1_AgeatEyeOpening_d, X8.1_AdultForearmLen_mm, X10.1_PopulationGrpSize))


#remove traits without any information gain (only consist of one value)
dataset <- subset(dataset, select = -c(Motility, ParentalCare, X12.2_Terrestriality))


#remove traits that are almost identical to other traits
dataset <- subset(dataset, select = -c(PullStrength, NumOffspring, BreedingInterval, 
                                       Diet, AVGWeight, MaturityReachFemale,
                                       MaturityReachMale, X22.2_HomeRange_Indiv_km2,
                                       X5.3_NeonateBodyMass_g,X16.1_LittersPerYear,X7.1_DispersalAge_d))
```



# Model Selection
The Generalized Linear Model (GLM) must be made. The Domestication column is used as the dependent 
variable, and the other columns are the predictor variables. For the model selection the phylolm package 
and phyloglmstep function are used.


### Load packages
```{r packages}
library(phylolm)
source("https://raw.githubusercontent.com/naturalis/angiosperm-traits/master/phyloglmstep.R")
```


### Data

A dataset without any missing values is made. It contains the Domestication trait, and also the 
CanonicalNames to match the data with the tree. The tree tips that aren't in the final data, are dropped.

```{r data}
# Complete trait set, with Domestication and CanonicalNames
columns <- dataset[,c(1,7:78)]

# The predictor variables from the VIF analysis + Domestication and CanonicalNames
data <- columns[,c(1,2,3,4,9,14,16,24,25,27,31,32,33,35,52)]

# Only the rows without any missing values are collected
data <- data[complete.cases(data),]

# Dropping species from the tree, to make the amount of species the same in the data and tree
#setdiff(data$CanonicalName, tree$tip.label)
dropTips <- setdiff(tree$tip.label, data$CanonicalName)
phyloTree <- drop.tip(tree, dropTips)

# Domesticated species
#domData <- dataset[dataset$Domestication == 1,]
#domData <- domData[,c(1,7,8,9,14,19,21,29,30,32,36,37,38,40,57)]

rm(columns, dropTips)
```


### Binary dependent variable

The dependent variable should be in a binary state. This means that the Domestication trait needs to be
converted to 0-1 instead of the 1-2 it's now. The 'wild' state (2) will be converted to the zero state (0). This way 'wild' appears as 0 and 'domesticated' appears as 1.

```{r zero state}
data$Domestication[data$Domestication==2] <- 0 
```

### Formulas

The input formulas are defined, which are going to be used as input for the phyloglmstep function.

```{r declare formulas}
# Lowest VIF values - with all traits
formula1 <- Domestication ~ X1.1_ActivityCycle + X5.1_AdultBodyMass_g + X12.1_HabitatBreadth + X21.1_PopulationDensity_n_km2 + X10.2_SocialGrpSize + MatingSystem + YearRoundBreeding + Horns_Antlers + AVGMovingSpeed + AVGTravelDistance + Aspect + ClayPercentage + OrganicCarbon

# Without X12.1_HabitatBreadth
formula2 <- Domestication ~ X1.1_ActivityCycle + X5.1_AdultBodyMass_g + X21.1_PopulationDensity_n_km2 + X10.2_SocialGrpSize + MatingSystem + YearRoundBreeding + Horns_Antlers + AVGMovingSpeed + AVGTravelDistance + Aspect + ClayPercentage + OrganicCarbon


```


### Model selection

With the use of the phyloglmstep() function, written by Rutger Vos.

```{r model selection}
phyloglmstep(formula2, starting.formula = NULL, data=data, phy=phyloTree, 
             method="logistic_MPLE", direction = "forward", trace = 2, 
             btol = 10, log.alpha.bound = 4, start.beta=NULL, 
             start.alpha=NULL, boot = 0, full.matrix = TRUE, k=2)
```


