---
title: "Correlates of domesticability in ungulates"
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE, message=FALSE}

# To produce an appealing output document, we will hide the code from it, hence
# echo = FALSE, and same for warnings:

knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

# Load the required packages:

library(ape)
library(dplyr)
library(usdm)
library(caret)
library(corrplot)
library(phylolm)
library(performance)
library(see)
library(gridExtra)
library(tidyverse)
library(naniar)
library(picante)
library(mice)
```

# Introduction

All of the large animals that humans have ever domesticated and that went on to
have a significant impact on the fates of human societies are hooved mammals.
Assuming that humans did not simply domesticate animal species at random (a safe
assumption given how taxonomically restricted the domesticates are), there must
have been traits that make some species attractive for domestication, and others
less so. In other words, there most be correlates of domesticability. In this
study, we seek to identify what these are.

# Materials and methods

## Data collection

Many scientists have discussed what trait values are required for an animal to
be amenable for domestication. For example, in the influential popular science 
book _Guns, Germs, and Steel_, author Jared Diamond {ref} enumerates a number of 
behavioral, ecological, and life history traits about which he asserts that each
has to fall into place or domestication will fail (he calls this the Karenina 
principle). Other authors {ref, ref} have proposed other traits, so there is no 
definitive list. Hence, here we will cast a wide net in an approach where we 
will test which combination of traits best predicts the domesticated state 
across the ungulates.

### Morphological, life history, and behavioral traits

Ungulates were initially domesticated for food and some of their other body
parts as raw materials, e.g. hides, horns, bones, sinew, etc. Their usefulness
as transport and draft animals and as providers of milk and manure came later
(cf. the 'secondary products revolution' {ref}). As such, early domestication 
was a kind of replacement of hunting, which is a lot of work and potentially 
dangerous. Given the expended effort, optimal foraging theory suggests that
hunting (or, subsequently, domesticating) small ungulates such as chevrotains is
relatively unattractive. Conversely, very large ungulates such as rhinoceroses 
are too dangerous. This suggest that body mass by itself might play a role, as 
might many of the traits that allometrically scale with it. For example, Diamond 
argues that long gestation lengths and life spans - which scale with body mass
and size - are unattractive. For these kinds of traits we combine data
from the PanTHERIA database of mammal diversity {ref} with additional literature
study of other plausible candidate traits.

**Add a section here about the traits that you and Zoe collected: what are they,
why these ones, where do they come from**

```{r morpho}
ungulatesData <- read.csv("https://raw.githubusercontent.com/naturalis/trait-organismal-ungulates/master/data/CSV/ungulatesTraits.csv")
```

### Abiotic niche dimensions

None of the domesticated ungulates are from tropical rain forests. Multiple 
explanations are possible for this. For example, grassland biomes select for
large social group (herd) formation, which in turn may be attractive. Or, 
provisioning rain forest animals with the food they require is cumbersome - or
indeed anything else about the habitat they require is hard to reproduce
sufficiently. In any case, the biotope and its properties can be hypothesized to
play a role. To derive these properties, we collected geo-referenced occurrences 
for the ungulates from GBIF {ref}, and used their latitude/longitude coordinates
to obtain pixel values for 42 GIS layers from selected data sources {ref, ref, 
ref, ref}. We then applied an outlying mean index (OMI) transformation to these 
raw pixel values to obtain single point estimates for each GIS layer, for each 
species. Simply put, this gives us an estimate of, for example, annual average 
temperature in the localities where occurrences of zebras have been recorded.


```{r abio}
omi <- read.csv("https://raw.githubusercontent.com/naturalis/trait-geo-diverse-ungulates/master/results/OMI/niche_traits.csv")
```

### Taxonomy and phylogeny

In this study, we combine species data from multiple sources. This includes the 
data we have described in the previous sections but also phylogeny. As our study
involves the comparative analysis of biological species, which are always 
related by shared descent however distant, we must account for 
pseudoreplication by incorporating phylogeny, for which we use a previously 
published supertree {ref}. We normalize all these different sources on a single 
taxonomy {ref} as the authoritative source for canonical names. Mammalian 
taxonomy is relatively well-resolved, but we did have to make a few minor 
decisions. As an initial remark, the name 'ungulates' is folk taxonomy, the
current systematics of our study species defines them as the terrestrial members
of the Cetartiodactyla and the Perissodactyla, which jointly form a paraphyletic
clade that excludes the whales. Subsequently, a taxonomic rule in zoological 
nomenclature that is only variably observed, is that domesticates are not 
assigned a distinct specific epithet (only subspecific ones). This manifested in
the phylogeny, which includes the tip _Equus asinus_ - domesticated donkeys. We
relabeled this as the wild ancestor _Equus africanus_. (This has no effect on the
rest of the analysis, as the patristic distance between either of these and the
other species in the data is identical.)

```{r merge data}
# join on canonical name
names(omi)[names(omi)=="X"] <- "CanonicalName"
dataset <- merge(ungulatesData, omi, by="CanonicalName")

# substitute spaces with underscores and use as row names to match the tree
row.names(dataset) <- gsub(" ", "_", dataset$CanonicalName)

# read tree, correct donkey name
tree <- read.tree("https://raw.githubusercontent.com/naturalis/trait-organismal-ungulates/master/data/phylogeny/ungulates.tree")
tree$tip.label[tree$tip.label=="Equus_asinus"] <- "Equus_africanus"

# match tree and data
res <- match.phylo.data(tree, dataset)
tree <- res[1]$phy
dataset <- res[2]$data

rm(ungulatesData, omi, res)
```

### Data filtering

For the success of the subsequent analysis, some data columns need to be 
omitted. Firstly, some of the data was imported from the PanTHERIA data set, 
which spans all mammals but for which some of the columns were invariant for our 
study species. This was the case for four columns. We also removed
the 13 columns with more than 50% missing data. Lastly, some of our columns are
(near) copies of other columns. For example, we searched for data on the pull 
strength and carry weight of ungulates and found that a common rule is that 
these are scaled versions of the body mass. Such collinearity poses problems
in some of our calculations, so we removed the three columns where that is the
case.

```{r filter}
# remove traits without any information gain (only consist of one value)
dataset <- subset(dataset, select = -c(nearZeroVar(dataset)))

# remove columns with >50% NA
dataset <- dataset[, which(colMeans(!is.na(dataset)) > 0.5)]

# remove traits that are almost identical to other traits
# XXX Why are we removing neonate body mass? I reverted that
# XXX Why do we have both X15.1_LitterSize AND NumOffspring?
# XXX What's the difference between LifeSpan and X17.1_MaxLongevity_m
dataset <- subset(dataset, select = -c(AVGBodyMass, CarryWeight, PullStrength, NumOffspring))
```

After the data merging and filtering we now have a data set of 74 variables, of
which 14 are categorical/factors and 60 are numeric, for 148 species.

```{r preprocess}
# remove columns that contain non-data taxonomic information
dataset <- subset(dataset, select = -c(CanonicalName, BinomialName, Order, Family, Genus, Species, ID_EoL))

# make sure the columns are typed correctly. these are factors, the rest is numeric
factors <- c(
  'Domestication','X1.1_ActivityCycle','X6.1_DietBreadth','X12.1_HabitatBreadth',
  'X6.2_TrophicLevel','Diet','Sociality','SocialHierarchy','NumMales',
  'MatingSystem','YearRoundBreeding','DevelopmentStrategy','HeadOrnaments',
  'NaturalPredators'
)

for (colname in names(dataset)) {
  if ( colname %in% factors ) {
    dataset[[colname]] <- as.factor(dataset[[colname]])
  }
  else {
    dataset[[colname]] <- as.numeric(dataset[[colname]])
  }
}
rm(colname, factors)
```

## Variable selection

The variables we have combined in our data set show multicollinearity in a 
number of clusters (Supplementary Figure 1). Among these clusters are those 
where all variables are affected by temperate, such as bio1 through bio11 from 
BIOCLIM {ref} as well as ENVIREM {ref} temperature-related variables, a similar 
cluster of variables related to precipitation in both bioclimatic data sets, a 
cluster of variables that all scale with body mass, and other smaller clusters. 
We have to address this multicollinearity as it may make the model behave 
erratically, affect the estimates of coefficients, and the significance. 

```{r corrplot, results='markup'}
# take only the numeric predictors and remove the dependent variable
startPredictors <- select_if(dataset, is.numeric)

# the first correlation matrix is a visualization of the dataset without the 
# removal of any traits. There is a lot of collinearity present.
matrix <- cor(startPredictors, use = "pairwise.complete.obs")
corrplot(matrix, type="full", order = "hclust", tl.pos = "lt", tl.col = "black", 
         tl.cex = 0.4, diag=F, title="Supplementary Figure 1")

```

Multicollinearity can be addressed among other ways by removing variables with 
high pairwise correlation coefficients, or, more directed, whose Variable 
Inflation Factor (VIF) is greater than 10 {ref}. However, if this is the 
sole criterion for including or excluding variables at this stage, we might be
removing strong predictors. Removing more, but weaker, predictors at this stage
might result in a final model that performs better. Hence, we might include in
our criteria for variable selection their individual fit in predicting the
dependent variable (i.e. domesticability). We therefore performed a separate
PhyloGLM {ref} for each numeric predictor and stored the AIC of each of these
model fits as a criterion for ranking the predictors later.

```{r indiv_fit}
varnames <- names(startPredictors)
scores <- list()
for ( var in varnames ) {
  
  # this is how to create a formula object where a term is stored in a variable
  formula <- paste0('Domestication', ' ~ ', var)
  res <- phyloglm(formula, data = dataset, phy = tree, method = "logistic_MPLE")
  
  # populates a list of key/value pairs, each key is a predictor name
  scores[[ var ]] <- list(aic=res$aic, keep=TRUE)
}
rm(res, formula, var, varnames)
```

We then used the inverse of the absolute pairwise correlation coefficients for
all combinations of numeric predictors to obtain a distance matrix for 
clustering of the variables into islands of multicollinearity. To inspect the 
finer-grained patterns in dendrogram form, we performed hierarchical clustering 
with complete linkage (Supplementary Figure 2).

Because the dendrogram has 5 basal clades and because of the contours in the
correlation plot (Supplementary Figure 1) we created 5 centers by _k_-means
clustering. 

```{r cluster_vars}
# our distance matrix is the inverse of the absolute correlation coefficients
distmat <- 1 - abs(matrix)

# we use k-means clustering to get distinct 'islands'. based on the correlation
# matrix plot, 5 distinct clusters seems reasonable.
kmeans_res <- kmeans(as.dist(distmat), 5)
for ( var in names(scores) ) {
  scores[[var]]$cluster <- kmeans_res$cluster[[var]]
}

# this is basically just to produce the plot below
hclust_res <- hclust(as.dist(distmat), method = "complete")
plot(hclust_res, main = "Supplementary Figure 2")
rm(distmat, hclust_res, matrix, var, kmeans_res)
```

We then proceeded to reduce the variable set such that VIF<10 for all variables
with the following algorithm:

1. Compute the VIF. Stop if all VIF<10.
2. Iterate over all 5 clusters.
3. In each cluster, remove the variable with the lowest AIC if its VIF>10.
4. Go to 1.

This results in a remaining set of 17 variables.

```{r vifalgo}
while(TRUE) {
  is_complete <- TRUE # are-we-done-yet switch
  
  # compute the VIFs for all (current) variables and apply row names
  vif_raw <- vif(startPredictors)
  row.names(vif_raw) <- vif_raw$Variables
  
  # add/update VIFs in scores structure
  for ( var in names(scores) ) {
    scores[[var]]$VIF <- vif_raw[var,]$VIF
    if ( scores[[var]]$keep && scores[[var]]$VIF > 10 ) {
      is_complete <- FALSE # at least one var has VIF>10, not done yet
    }
  }
  
  # the switch wasn't flipped, we're done, exit the infinite loop
  if (is_complete ) {
    break
  }
  
  # iterate over clusters
  for ( cls in 1:5 ) {
    weakest_link <- NULL # lowest AIC variable in focal cluster
    for ( var in names(startPredictors) ) {
      s <- scores[[var]]
      
      # var is in cluster, is collinear, but not yet pruned
      if ( s$cluster == cls && s$VIF > 10 && s$keep ) {
        
        # we have no var yet, or focal one is worse than seen so far
        if ( is.null(weakest_link) || s$aic < scores[[weakest_link]]$aic ) {
          weakest_link <- var
        }
      }
    }
    if ( ! is.null(weakest_link) ) {
      scores[[weakest_link]]$keep <- FALSE
      startPredictors[[weakest_link]] <- NULL
    }
  }
}
rm(s, scores, vif_raw, cls, is_complete, var, weakest_link)
```

## Model Selection

We hypothesize that among the selected numeric variables and the factors is a 
combination that predicts domesticability in a phylogenetic generalized linear
model. To establish this combination we perform stepwise model selection.

```{r model_selection}
# add factor predictors to numeric predictors
factors <- select_if(dataset, is.factor)
selected_dataset <- as.data.frame(cbind(factors, startPredictors))
#rm(factors, startPredictors)

# impute missing values
# https://www.r-bloggers.com/2015/10/imputing-missing-data-with-r-mice-package/
imputed_dataset <- complete(mice(selected_dataset, m = 1, meth = 'pmm'), 1 )

# compose formula string
predictor_names <- names(selected_dataset[2:length(selected_dataset)])
terms <- paste(predictor_names, collapse = ' + ')
full_formula <- paste('Domestication', '~', terms)
#rm(terms, predictor_names)

res <- phyloglmstep(full_formula, data=imputed_dataset, phy=tree, method="logistic_IG10", direction="forward", btol=100, trace=2)
```


# 5 Modelling analysis
The phylogenetic generalized linear modelling analysis optimizes the model. The
function phyloglm from the phylolm package is used for this.
```{r model analysis, results='hide'}
# Converting dependent variable to binary state
# Domesticated = 1, wild = 0
dataset$Domestication[dataset$Domestication==2] <- 0

# Construct model
phyloGLM <- phyloglm(formula = finalFormula, data = dataset, phy = tree, method = "logistic_MPLE", btol = 36, log.alpha.bound = 4)
summary(phyloGLM)
```


# 6 Model evaluation
The model is evaluated to check whether the model constructed, is still in
compliance with the assumptions that have to be met when performing a GLM. The
package 'performance' is used for this. The function 'check_model()' is used for
a comprehensive model check. This function is not appropriate for models
including phylogeny. The check will still be done, but the resulting
interpretation may be skewed.
```{r model evaluation}
GLM <- glm(formula = finalFormula, data = dataset, family = gaussian)

check_model(GLM)
```


# 7 Plots
The results are visualized using different plots.

```{r plots}
source("Plots.R")

```
