---
title: "Correlates of domesticability in ungulates"
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE, message=FALSE}

# To produce an appealing output document, we will hide the code from it, hence
# echo = FALSE, and same for warnings:

knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

# Load the required packages:

library(ape)
library(dplyr)
library(usdm)
library(caret)
library(corrplot)
library(phylolm)
library(performance)
library(see)
library(gridExtra)
library(tidyverse)
library(naniar)
library(picante)
```

# Introduction

All of the large animals that humans have ever domesticated and that went on to
have a significant impact on the fates of human societies are hooved mammals.
Assuming that humans did not simply domesticate animal species at random (a safe
assumption given how taxonomically restricted the domesticates are), there must
have been traits that make some species attractive for domestication, and others
less so. In other words, there most be correlates of domesticability. In this
study, we seek to identify what these are.

# Materials and methods

## Ungulate traits and domestication

Many scientists have discussed what trait values are required for an animal to
be amenable for domestication. For example, in the influential popular science 
book _Guns, Germs, and Steel_, author Jared Diamond {ref} enumerates a number of 
behavioral, ecological, and life history traits about which he asserts that each
has to fall into place or domestication will fail (he calls this the Karenina 
principle). Other authors {ref, ref} have proposed other traits, so there is no 
definitive list. Hence, here we will cast a wide net in an approach where we 
will test which combination of traits best predicts the domesticated state 
across the ungulates.

### Morphological, life history, and behavioral traits

Ungulates were initially domesticated for food and some of their other body
parts as raw materials, e.g. hides, horns, bones, sinew, etc. Their usefulness
as transport and draft animals and as providers of milk and manure came later
(cf. the 'secondary products revolution' {ref}). As such, early domestication 
was a kind of replacement of hunting, which is a lot of work and potentially 
dangerous. Given the expended effort, optimal foraging theory suggests that
hunting (or, subsequently, domesticating) small ungulates such as chevrotains is
relatively unattractive. Conversely, very large ungulates such as rhinoceroses 
are too dangerous. This suggest that body mass by itself might play a role, as 
might many of the traits that allometrically scale with it. For example, Diamond 
argues that long gestation lengths and life spans - which scale with body mass
and size - are unattractive. For these kinds of traits we combine data
from the PanTHERIA database of mammal diversity {ref} with additional literature
study of other plausible candidate traits.

**Add a section here about the traits that you and Zoe collected: what are they,
why these ones, where do they come from**

```{r morpho}
ungulatesData <- read.csv("https://raw.githubusercontent.com/naturalis/trait-organismal-ungulates/master/data/CSV/ungulatesTraits.csv")
```

### Abiotic niche dimensions

None of the domesticated ungulates are from tropical rain forests. Multiple 
explanations are possible for this. For example, grassland biomes select for
large social group (herd) formation, which in turn may be attractive. Or, 
provisioning rain forest animals with the food they require is cumbersome - or
indeed anything else about the habitat they require is hard to reproduce
sufficiently. In any case, the biotope and its properties can be hypothesized to
play a role. To derive these properties, we collected geo-referenced occurrences 
for the ungulates from GBIF {ref}, and used their latitude/longitude coordinates
to obtain pixel values for 42 GIS layers from selected data sources {ref, ref, 
ref, ref}. We then applied an outlying mean index (OMI) transformation to these 
raw pixel values to obtain single point estimates for each GIS layer, for each 
species. Simply put, this gives us an estimate of, for example, annual average 
temperature in the localities where occurrences of zebras have been recorded.


```{r abio}
omi <- read.csv("https://raw.githubusercontent.com/naturalis/trait-geo-diverse-ungulates/master/results/OMI/niche_traits.csv")
```

### Taxonomy and phylogeny

In this study, we combine species data from multiple sources. This includes the 
data we have described in the previous sections but also phylogeny. As our study
involves the comparative analysis of biological species, which are always 
related by shared descent however distant, we must account for 
pseudoreplication by incorporating phylogeny, for which we use a previously 
published supertree {ref}. We normalize all these different sources on a single 
taxonomy {ref} as the authoritative source for canonical names. Mammalian 
taxonomy is relatively well-resolved, but we did have to make a few minor 
decisions. As an initial remark, the name 'ungulates' is folk taxonomy, the
current systematics of our study species defines them as the terrestrial members
of the Cetartiodactyla and the Perissodactyla, which jointly form a paraphyletic
clade that excludes the whales. Subsequently, a taxonomic rule in zoological 
nomenclature that is only variably observed, is that domesticates are not 
assigned a distinct specific epithet (only subspecific ones). This manifested in
the phylogeny, which includes the tip _Equus asinus_ - domesticated donkeys. We
relabeled this as the wild ancestor _Equus africanus_. (This has no effect on the
rest of the analysis, as the patristic distance between either of these and the
other species in the data is identical.)

```{r merge data}
# join on canonical name
names(omi)[names(omi)=="X"] <- "CanonicalName"
dataset <- merge(ungulatesData, omi, by="CanonicalName")

# substitute spaces with underscores and use as row names to match the tree
row.names(dataset) <- gsub(" ", "_", dataset$CanonicalName)

# read tree, correct donkey name
tree <- read.tree("https://raw.githubusercontent.com/naturalis/trait-organismal-ungulates/master/data/phylogeny/ungulates.tree")
tree$tip.label[tree$tip.label=="Equus_asinus"] <- "Equus_africanus"

# match tree and data
res <- match.phylo.data(tree, dataset)
tree <- res[1]
dataset <- res[2]$data

rm(ungulatesData, omi, res)
```

### Filtering incomplete data

The traits that consist of more than 100 missing values, traits that have no
information gain and traits that are almost identical to other traits are
removed.
```{r misc preprocessing}
# Rename column with dots in the name
names(dataset)[names(dataset)=="X21.1_PopulationDensity_n.km2"] <- "X21.1_PopulationDensity_n_km2"

# Remove traits that contain >50% missing variables
# Check the number and % missing values for all variables in the dataset 
miss_var_summary(dataset) 


## Remove columns with >50% NA
# let's experiment with the 0.5 threshold, might be changed
dataset <- dataset[, which(colMeans(!is.na(dataset)) > 0.5)]

# Remove traits without any information gain (only consist of one value)
#Identify which variables have zero variance
nearZeroVar(dataset)

dataset <- subset(dataset, select = -c(nearZeroVar(dataset)))

# Remove traits that are almost identical to other traits
# XXX Why are we removing neonate body mass
dataset <- subset(dataset, select = -c(AVGBodyMass, CarryWeight, PullStrength, X5.3_NeonateBodyMass_g))

#Remove columns that contain non-data taxonomic information
dataset <- subset(dataset, select = -c(CanonicalName, BinomialName, Order, Family, Genus, Species, ID_EoL))
```

### Set as factors
Some of the columns are not seen as categorical, but as numeric. These must be converted to factors.
```{r as factor}

#TODO: Check if all columns are represented 
dataset$Domestication <- as.factor(dataset$Domestication)
dataset$X1.1_ActivityCycle <- as.factor(dataset$X1.1_ActivityCycle)
dataset$X6.1_DietBreadth <- as.factor(dataset$X6.1_DietBreadth)
dataset$X12.1_HabitatBreadth <- as.factor(dataset$X12.1_HabitatBreadth)
dataset$X6.2_TrophicLevel <- as.factor(dataset$X6.2_TrophicLevel)
dataset$Diet <- as.factor(dataset$Diet)
dataset$Sociality <- as.factor(dataset$Sociality)
dataset$SocialHierarchy <- as.factor(dataset$SocialHierarchy)
dataset$NumMales <- as.factor(dataset$NumMales)
dataset$MatingSystem <- as.factor(dataset$MatingSystem)
dataset$YearRoundBreeding <- as.factor(dataset$YearRoundBreeding)
dataset$DevelopmentStrategy <- as.factor(dataset$DevelopmentStrategy)
dataset$HeadOrnaments <- as.factor(dataset$HeadOrnaments)
dataset$NaturalPredators <- as.factor(dataset$NaturalPredators)

```

# 3 VIF-Analysis
There is probably collinearity present amongst the traits in the dataset.
Collinearity can lead to bias in the model, so we must correct for this. This can be done by running a variable inflation factor analysis (VIF). This script only requires the 'dataset' variable and after it is run, it will output the 'predictors' variable.
```{r corrplot, results='markup'}
# Do the VIF analysis
## BEGIN VALUES
# Take only the numeric predictors and remove the dependent variable
startPredictors <- dplyr::select_if(dataset, is.numeric)
startPredictors$Domestication <- NULL

# The first correlation matrix is a visualization of the dataset without the removal of any traits. There is a lot of collinearity present.
matrix <- cor(startPredictors, use = "pairwise.complete.obs")
corrplot(matrix, type="lower", order = "hclust", tl.pos = "l", tl.col = "black", tl.cex = 0.4)

```

```{r VIF, results='markup'}
## CHECK CORRELATION MATRIX
# The corrplot package is used to visualize the collinearity between the traits,
# using the correlation matrix. 
# With the findCorrelation() function, 
# the columns with the highest collinearity are found. 
# By visualizing the matrix, the outcome of the findCorrelation function can be 
# easily checked. The VIF function can be run to easily check if the results are
# still infinite. A cutoff value of .83 is chosen, using trial and error.
matrix <- cor(startPredictors, use = "pairwise.complete.obs")
delColumn <- findCorrelation(matrix, cutoff = .83, verbose = FALSE, exact=FALSE)
startPredictors <- startPredictors[,-c(delColumn)]



vif(startPredictors)


```
#Precipitation
explanation
```




```

#Temperature
explanation
```





```
#BodyMass
explanation


## INSERT ADULTBODYMASS
# The set of predictors is declared for the VIF analysis, and a last check for collinearity using the visualized matrix is done. During the first round, AdultBodyMass was removed due to a high collinearity with other traits. However, because Adultbodymass is hypothesized to be a biologically relevant trait, it will be reinserted in this step.
vifPredictors <- cbind(dataset$X5.1_AdultBodyMass_g, startPredictors)
colnames(vifPredictors)[1] <- "X5.1_AdultBodyMass_g"

# Check VIF -> Adultbodymass is highly correlated with the other traits.
vif(vifPredictors)

# CarryWeight has a high collinearity with AdultBodyMass (both inf). Since AdultBodyMass
#  is hypothesized to be biologically relevant, CarryWeight will be removed.
vifPredictors$CarryWeight <- NULL
vif(vifPredictors)

# Check correlation matrix
# A visualization of the dataset after the cutoff value is implemented and 
# AdultBodyMass is added.
matrix <- cor(vifPredictors, use = "pairwise.complete.obs")
corrplot(matrix, type="lower", order = "hclust", tl.pos = "l", tl.col = "black", tl.cex = 0.7)


## VIF ANALYSIS
# Check VIF -> PETWarmestQuarter has highest VIF value above 10
# Remove PETWarmestQuarter
vif(vifPredictors)
vifPredictors$PETWarmestQuarter <- NULL

# Check VIF -> bio3 has highest VIF value above 10
# Remove bio3
vif(vifPredictors)
vifPredictors$bio3 <- NULL

# Check VIF -> bio2 has highest VIF value above 10
# Remove bio2
vif(vifPredictors)
vifPredictors$bio2 <- NULL

# Check VIF -> bio14 has highest VIF value above 10
# Remove bio14
vif(vifPredictors)
vifPredictors$bio14 <- NULL

# Check VIF -> PETDriestQuarter has highest VIF value above 10
vif(vifPredictors)
vifPredictors$PETDriestQuarter <- NULL

# Check VIF -> AVGFoodConsumption has highest VIF value above 10
vif(vifPredictors)
vifPredictors$AVGFoodConsumption <- NULL

# Check VIF -> X13.1_AdultHeadBodyLen_mm has highest VIF value above 10
vif(vifPredictors)
vifPredictors$X13.1_AdultHeadBodyLen_mm <- NULL

# Check VIF -> All VIF values under 10 (except AdultBodyMass, which is 11)
vif(vifPredictors)


## FINAL PREDICTOR SET
# Final set of numeric predictors, containing:  X5.1_AdultBodyMass_g, X9.1_GestationLen_d,
# X15.1_LitterSize, X17.1_MaxLongevity_m, X21.1_PopulationDensity_n_km2, 
# X23.1_SexualMaturityAge_d, X10.2_SocialGrpSize, X25.1_WeaningAge_d, Lifespan,
# AVGMovingSpeed, AVGTravelDistance, Aspect, BulkDensity, ClayPercentage, minTempWarmest,
# PETseasonality, PETWettestQuarter, OrganicCarbon, PhCaCL, Slope, bio15, bio18, bio19
vifPredictors

## FINAL VALUES
# The final correlation matrix is the final product after removing all highly correlated
# traits. 
matrix <- cor(vifPredictors, use = "pairwise.complete.obs")
corrplot(matrix, type="lower", order = "hclust", tl.pos = "l", tl.col = "black", tl.cex = 0.7)

## REMOVE VARIABLES
rm(startPredictors, matrix, delColumn)

# Final VIF values
vif(vifPredictors)
```

# 4 Model Selection
A Generalized Linear Model (GLM) must be made. The Domestication column is
used as the dependent variable, and the other columns are the predictor
variables. For the model selection the phylolm package and phyloglmstep function
are required.
```{r model selection, results='hide'}
# Add factor predictors to numeric predictors
factors <- dplyr::select_if(dataset, is.factor)
predictors <- cbind(vifPredictors, factors)

# Do the model selection
source("ModelSelection.R")

# Remove unneeded variables
rm(vifPredictors, factors)
```


# 5 Modelling analysis
The phylogenetic generalized linear modelling analysis optimizes the model. The
function phyloglm from the phylolm package is used for this.
```{r model analysis, results='hide'}
# Converting dependent variable to binary state
# Domesticated = 1, wild = 0
dataset$Domestication[dataset$Domestication==2] <- 0

# Construct model
phyloGLM <- phyloglm(formula = finalFormula, data = dataset, phy = tree, method = "logistic_MPLE", btol = 36, log.alpha.bound = 4)
summary(phyloGLM)
```


# 6 Model evaluation
The model is evaluated to check whether the model constructed, is still in
compliance with the assumptions that have to be met when performing a GLM. The
package 'performance' is used for this. The function 'check_model()' is used for
a comprehensive model check. This function is not appropriate for models
including phylogeny. The check will still be done, but the resulting
interpretation may be skewed.
```{r model evaluation}
GLM <- glm(formula = finalFormula, data = dataset, family = gaussian)

check_model(GLM)
```


# 7 Plots
The results are visualized using different plots.

```{r plots}
source("Plots.R")

```