---
title: "Pipeline_Analysis"
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# 1 Import data
Our first step is to import packages and data into R. The trait data and OMI data also need to be merged.

### Load packages
```{r Load packages, message=FALSE}
library(ape)
library(dplyr)
library(usdm)
library(caret)
library(corrplot)
library(phylolm)
library(performance)
library(see)
library(gridExtra)
```

### Import data
The datasets can be found in different GitHub repositories. The ungulate
dataset and tree can be found in the trait-organismal-ungulates repository. 
The OMI data is found in the trait-geo-diverse-ungulates repository.
```{r import data}
ungulatesData <- read.csv("https://raw.githubusercontent.com/naturalis/trait-organismal-ungulates/master/data/CSV/ungulatesTraits.csv")
omi <- read.csv("https://raw.githubusercontent.com/naturalis/trait-geo-diverse-ungulates/master/results/OMI/niche_traits.csv")
tree <- read.tree("https://raw.githubusercontent.com/naturalis/trait-organismal-ungulates/master/data/phylogeny/ungulates.tree")
```

### Merge datasets
The ungulate data and OMI data have to be merged into one dataset. The EoL-ID is
removed and the data is merged by the canonical name (present in both the
datasets). The last step is to replace the spaces in the canonical name with
underscores, to match the species names in the tree.
```{r merge data}
ungulatesData <- ungulatesData[2:60]
names(omi)[names(omi)=="X"] <- "CanonicalName"
dataset <- merge(ungulatesData, omi, by="CanonicalName")
dataset$CanonicalName <- gsub(" ", "_", dataset$CanonicalName)

# Clean up the global environment 
rm(ungulatesData, omi)
```

# 2 Preprocessing
### Equalize species in tree and dataset
To start, 'Equus asinus' is renamed to the 'Equus africanus' in the tree, to
match the dataset. The species that aren't in the tree are dropped from the
dataset. The species that aren't in the dataset are dropped from the tree.
```{r check if names match between dataset and tree, results='hide'}
# Changed Equus asinus to Equus africanus in the tree 
tree$tip.label[tree$tip.label=="Equus_asinus"] <- "Equus_africanus"

# Check Which species aren't in the tree
dropRows <- setdiff(dataset$CanonicalName, tree$tip.label)

# Drop rows that aren't in the tree (check manually if these are domesticated)
row.names(dataset) <- dataset$CanonicalName
dataset <- dataset[!(row.names(dataset) %in% dropRows), ]

# Drop tips that aren't in dataset
dropTips <- setdiff(tree$tip.label, dataset$CanonicalName)
tree <- drop.tip(tree, dropTips)

# Final check to see if there are any differences
setdiff(dataset$CanonicalName, tree$tip.label)
setdiff(tree$tip.label, dataset$CanonicalName)

rm(dropRows, dropTips)
```

### Miscellaneous preprocessing
The dots in the column names are replaced with underscores. After that, the
traits that consist of more than 100 missing values, traits that have no
information gain and traits that are almost identical to other traits are
removed.
```{r misc preprocessing}
# Rename column with dots in the name
names(dataset)[names(dataset)=="X21.1_PopulationDensity_n.km2"] <- "X21.1_PopulationDensity_n_km2"

# Remove traits that (almost) only consist of missing values (>100 NA)
dataset <- subset(dataset, select = -c(X5.4_WeaningBodyMass_g, X13.3_WeaningHeadBodyLen_mm, X13.2_NeonateHeadBodyLen_mm, X18.1_BasalMetRate_mLO2hr, X5.2_BasalMetRateMass_g, X2.1_AgeatEyeOpening_d, X8.1_AdultForearmLen_mm, X10.1_PopulationGrpSize))

# Remove traits without any information gain (only consist of one value)
dataset <- subset(dataset, select = -c(Motility, ParentalCare, X12.2_Terrestriality))

# Remove traits that are almost identical to other traits
dataset <- subset(dataset, select = -c(PullStrength, NumOffspring, BreedingInterval, Diet, AVGWeight, MaturityReachFemale, MaturityReachMale, X22.2_HomeRange_Indiv_km2, X5.3_NeonateBodyMass_g, X16.1_LittersPerYear,X7.1_DispersalAge_d))
```

### Set as factors
Some of the columns are not seen as categorical, but as numeric. These must be converted
to factors.
```{r as factor}
dataset$X1.1_ActivityCycle <- as.factor(dataset$X1.1_ActivityCycle)
dataset$X6.1_DietBreadth <- as.factor(dataset$X6.1_DietBreadth)
dataset$X12.1_HabitatBreadth <- as.factor(dataset$X12.1_HabitatBreadth)
dataset$X6.2_TrophicLevel <- as.factor(dataset$X6.2_TrophicLevel)
dataset$Sociality <- as.factor(dataset$Sociality)
dataset$SocialHierarchy <- as.factor(dataset$SocialHierarchy)
dataset$NumMales <- as.factor(dataset$NumMales)
dataset$MatingSystem <- as.factor(dataset$MatingSystem)
dataset$YearRoundBreeding <- as.factor(dataset$YearRoundBreeding)
dataset$DevelopmentStrategy <- as.factor(dataset$DevelopmentStrategy)
dataset$HeadOrnaments <- as.factor(dataset$HeadOrnaments)
dataset$NaturalPredators <- as.factor(dataset$NaturalPredators)
```

# 3 VIF-Analysis
There is probably collinearity present amongst the traits in the dataset.
Collinearity can lead to bias in the model, so we must correct for this. This
can be done by running a variable inflation factor analysis (VIF). The VIF.R
script contains the whole VIF analysis. The R script is sourced below. This
script only requires the 'dataset' variable and after it  is run, it will output the 'predictors' variable.
```{r VIF, results='markup'}
# Do the VIF analysis
source("VIF.R")

# Final VIF values
vif(vifPredictors)
```

# 4 Model Selection
A Generalized Linear Model (GLM) must be made. The Domestication column is
used as the dependent variable, and the other columns are the predictor
variables. For the model selection the phylolm package and phyloglmstep function
are required.
```{r model selection, results='hide'}
# Add factor predictors to numeric predictors
factors <- dplyr::select_if(dataset, is.factor)
predictors <- cbind(vifPredictors, factors)

# Do the model selection
source("ModelSelection.R")

# Remove unneeded variables
rm(vifPredictors, factors)
```

#### Results:
      
       AIC     logLik Pen.logLik 
    28.346     -7.173     11.847 
    
    Parameter estimate(s) from MPLE:
    alpha: 0.01145475 
    
    Coefficients:
         (Intercept) X5.1_AdultBodyMass_g  DevelopmentStrategy        Horns_Antlers 
    -23.861324750517       0.000006403856       4.695920350903       8.733986057383 
      AVGMovingSpeed    AVGTravelDistance 
     -0.271395434194       0.006078751863 


# 5 Modelling analysis
The phylogenetic generalized linear modelling analysis optimizes the model. The
function phyloglm from the phylolm package is used for this.
```{r model analysis, results='hide'}
# Converting dependent variable to binary state
# Domesticated = 1, wild = 0
dataset$Domestication[dataset$Domestication==2] <- 0

# Construct model
phyloGLM <- phyloglm(formula = finalFormula, data = dataset, phy = tree, method = "logistic_MPLE", btol = 36, log.alpha.bound = 4)
summary(phyloGLM)
```

#### Results:

    
    Coefficients:
                              Estimate        StdErr z.value  p.value   
    (Intercept)          -9.3503949900  5.1361469886 -1.8205 0.068682 . 
    X5.1_AdultBodyMass_g  0.0000045743  0.0000020387  2.2437 0.024850 * 
    DevelopmentStrategy   1.9876564874  1.1964451827  1.6613 0.096653 . 
    Horns_Antlers         4.9891203727  1.8993795774  2.6267 0.008621 **
    AVGMovingSpeed       -0.2389830249  0.0950728567 -2.5137 0.011948 * 
    AVGTravelDistance     0.0046891017  0.0033020300  1.4201 0.155588  


# 6 Model evaluation
The model is evaluated to check whether the model constructed, is still in
compliance with the assumptions that have to be met when performing a GLM. The
package 'performance' is used for this. The function 'check_model()' is used for
a comprehensive model check. This function is not appropriate for models
including phylogeny. The check will still be done, but the resulting
interpretation may be skewed.
```{r model evaluation}
GLM <- glm(formula = finalFormula, data = dataset, family = gaussian)

check_model(GLM)
```


# 7 Plots
The results are visualized using different plots.

```{r plots}
source("Plots.R")

```